Are Delayed Issues Harder to Resolve?

### 2015-12-31 phaseDelay4Emse.tex
Updates from ShareLaTeX

\documentclass{sig-alternate}
\usepackage{multirow}
\usepackage[usenames, dvipsnames]{color}
\usepackage{colortbl}

@@

\usepackage[shortlabels]{enumitem} 
%\usepackage{times}

@@

\begin{document}

\conferenceinfo{ICSE}{'16 Austin, Texas}
\title{Live and Let Die? (Delayed Issues   not  Harder to Resolve)}
\numberofauthors{3}
\author{
%\alignauthor 
Tim Menzies \\
       \affaddr{Computer Science}\\
       \affaddr{NC State University, USA}\\
       \email{tim.menzies@gmail.com}
\and%alignauthor 
William Nichols, Forrest Shull \\
        \affaddr{Software Engineering Institute}\\
        \affaddr{Carnegie Mellon U, USA}\\
        \email{\{wrn,fjshull\}@sei.cmu.edu} 
\and %\alignauthor 
Lucas Layman \\
       \affaddr{Fraunhofer CESE} \\
       \affaddr{College Park, USA}\\ 
       \email{llayman@fc-md.umd.edu}
} 
\maketitle
\begin{abstract}
Many  practitioners and academics
believe in a delayed issue effect (DIE); i.e.
 as issues linger longer in a system,
 they become exponentially harder to resolve.
This belief
is often  used to justify 
major investments in  new development
processes that promise to retire more issues, sooner.

This paper tests for the delayed issue effect in
171 software projects conducted around the world in the period from 2006--2014.
To the best of our knowledge,  this is the largest study
yet published on this effect.
In our sample, we found no evidence for the  delayed issue effect; i.e.
the  time  to resolve 
issues in a later phase was not consistently  more than  
when  issues were resolved soon after their introduction.

This  result begs the question: how many other long-held
beliefs in software engineering are
unsupported by current data? 
\end{abstract}

%XXXwhat it this is due to TSP- 
% A category with the (minimum) three required fields
\vspace{1mm}
\noindent
{\bf Categories/Subject Descriptors:} 
D.2.8 [Software Engineering]: Process metrics.

@@

 Basili \& Boehm comment that since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[angle=270,width=3in]{img/b81.pdf}
 \end{center}
 \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}
Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.  McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this effect, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a proven
fact.
For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

% \begin{figure}[!t]
% \scriptsize
% \begin{center}
% \begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
%           &        & \multicolumn{2}{c|}{Time to}  &\\
%           &        & \multicolumn{2}{c|}{resolve} &\\
% Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
% Phase $i$&Phase $i$ & 1 & 1 & 1 & **\\
% Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
% Phase $i$ &Phase $i+2$ & 1 & 4& 4 &******** \\
% Phase $i$ &Phase $i+3$ & 1 & 8 & 8 & ****************  
%   \end{tabular}\end{center}
%  \caption{Projects data consistent with
%  the phase delay effect of \fig{curve1}.}\label{fig:curve1a}
% \end{figure}

% \begin{figure}[!t]
% \scriptsize
% \begin{center}
% \begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
%           &        & \multicolumn{2}{c|}{Time to}  &\\
%           &        & \multicolumn{2}{c|}{resolve} &\\
% Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
% Phase $i$&Phase $i$ & 1 & 1 & 1 & ** \\
% Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
% Phase $i$ &Phase $i+2$ & 1 & 2& 2 & **** \\
% Phase $i$ &Phase $i+3$ & 1 & 2 & 2 & **** \\
%  Phase $i$  &Phase $i+4$ &1 & 2& 2 &**** 
%   \end{tabular}\end{center}
%  \caption{Projects data consistent with
%  the phase delay effect of \fig{curve2}.}\label{fig:curve2a}
% \end{figure}
This paper calls into question all the above claims about the delayed
issue effect. We suggest that the delayed issue effect might have been an dominant
effect decades ago,  but {\em not}  for $21^{\mathit{st}}$
software development. 
 The delayed issue effect was first reported in 1981 in a era of punch card programming
and non-interactive environments~\cite{Boehm81}. In the 21$^\mathit{st}$ century, we  program in 
interactive environments with higher-level languages and better source code control
tools. Such tools allow for the faster refactoring of existing
code-- in which case, 
managing the changes required to fix (say) an incorrect requirements assumption
is far less    onerous    than before.

@@

% the SE community needs to stop
% justifying changes to SE
% practices via the phase delay effect. Secondly,   myths like phase delay should not have lasted decades
% in the literature without  critical reassessment.  As a community, we need to reflect more on our
% research practices and on how (and when) we reflect and revisit the prominent hypotheses of our field.

@@

That said, there are many empirical  findings that 
raise doubts that   general laws of SE even exist:
\be
\item
Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
on the relation of OO measures to defects.  Those results
 directly  contradict some of the laws listed by 
Endres \& Rombach~\cite{endres03}.
\item
Ray et al.~\cite{ray2014lang} tested if   strongly typed languages
predict for better code quality. In  728 projects,
they found  only a modest benefit in strong typing
(and caution that that effect may actually be due to other  conflating factors).
\item
Fenton \& Neil~\cite{fenton00,fenton00b}   critique the truism that
``pre-release fault rates for software
are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
amongst others). For the systems described in~\cite{fenton97}, they
show that software modules that were highly fault-prone
prior to release revealed very few faults after release.
\item
Meyer claims that   object-oriented (OO) encapsulation will
reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
that debugging an OO program is many times harder and
longer than debugging a standard procedural program~\cite{hatton98}.
\item
A truism of visual programming is that ``visual
representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
evidence for this claim is hardly conclusive~\cite{me00v}. 
\item
Numerous recent {\em local learning} results compare single models
learned from all available data to multiple models learned from clusters within the data~\cite{betten14,yang11,yang13,minku13,me12d,me11m,betta12,posnett11}.
A repeated result in those studies is that the local models generated the better effort
and defect predictions (better median results,
lower variance in the predictions).
\ee
% \item 
% Other work studies  found dozens of single
% models that performed worse that {\em committees of models}, all with % different opinions,
% that contribute to an {\em ensemble} prediction of software 
% development effort~\cite{kocaguneli2012value,azhar13}.
To be fair, 
SE is  not the only
field where practitioners hang on to persistent beliefs, even if the evidence
for those beliefs is not strong.
The medical profession applies  many practices based on studies
that have been disproved. For example,
a  recent article
in the Mayo Clinic Proceedings~\cite{prasad13} found  
146 medical practices based on studies 
in year $i$, but which were  reversed by subsequent trials within years $i+10$.
Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
Aschwanden warns that ``cognitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.

The cognitive issues that complicate medicine are also found in software engineering.
According to Passos et al.~\cite{passos11}, commercial developers
are all too willing to believe in general
truisms.
They  warn that developers
usually assume that the truisms they learn from a few past
projects are general to 
all their future projects. They comment ``past experiences were taken into account without 
much consideration for their context~\cite{passos11}.''  
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
  supposed software engineering    ``experts'' rarely use lessons
  from past projects to improve their future reasoning~\cite{jorgensen09}. 
 They note that
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
%% This leads to  detrimental
%% situations where groups or 
%% individuals working on the same project have conflicting beliefs, which they never debate
%or refine or unify.

@@

%\todo{\begin{itemize}
%\item Given the Intro and Motivation, this section should focus on the influence of phase delay in software engineering milieux
%\item Need to tone down the reliance on Beck. The important Beck point is that eliminating the cost-to-fix curve (resulting from phase delay) is one of the main goals of agile. His observations on new technologies are good too.
%\item Need to cite other popular sources of this information, e.g., McConnell's "Ounce of Prevention" article~\cite{mcconnell01}.
%\item Robert Glass's~\cite{glass02} book is very popular and contains an entry on this point
%\item Google Scholar Citations for phase delay works Boehm88~\cite{boehm88} (673), Boehm \& Basili~\cite{boehm01} (761), 
%\end{itemize}}

%XXX clumsy link here
%That said, where our results
%diverge from Beck is that we have observed this lack of phase delay in  projects that are not agile. Those
%projects are described below.

@@

\input{lucas}

@@

\input{lucasRelatedWork}

%\section{carterCharts}

%\input{carter}

\section{Delayed Issues not  Harder\newline to Resolve}

\input{data}

>>>>

\documentclass[smallcondesed]{svjour3}
\usepackage{multirow}
\usepackage[usenames, dvipsnames]{color}
\usepackage{colortbl}

@@

%\usepackage[shortlabels]{enumitem} 
%\usepackage{times}

@@

\pagenumbering{arabic} 
\begin{document}
\date{}
 \title{Are Delayed Issues Harder to Resolve?}

 \author{Tim Menzies, William Nichols, Forrest Shull, Lucas Layman}

\institute{
%\alignauthor 
Tim Menzies \at
       CS, North Carolina State University, USA,
                \email{tim.menzies@gmail.com} \
\and 
William Nichols, Forrest Shull \at 
        Software Engineering Institute ,
        Carnegie Mellon University, USA.
        \email{\{wrn,fjshull\}@sei.cmu.edu} 
\and  
Lucas Layman \at
        Fraunhofer CESE,  
        College Park, USA,
       \email{llayman@fc-md.umd.edu}
}
\maketitle
\begin{abstract}
Many  practitioners and academics
believe in a delayed issue effect (DIE); i.e.
 as issues linger longer in a system,
 they become very much harder to resolve.
This belief
is often  used to justify 
major investments in  new development
processes that promise to retire more issues, sooner.

This paper tests for the delayed issue effect in
171 software projects conducted around the world in the period from 2006--2014.
To the best of our knowledge,  this is the largest study
yet published on this effect.
We found no evidence for the  delayed issue effect; i.e.
the  time  to resolve 
issues in a later phase was  not consistently  more than  
when  issues were resolved soon after their introduction.

This paper documents the above studies and explores reasons for this  mismatch between this widely-held
belief and empirical data.
In  summary, DIE is not some constant across all projects. Rather, is may be an historical relic that only occurrs intermittently 
in  certain kinds of projects.  This is a significant result since it predicts that  new development
processes that promise to faster retire more issues will have only a  limited return on investment. 
\end{abstract}

 \vspace{1mm}
\noindent
{\bf Categories/Subject Descriptors:} 
D.2.8 [Software Engineering]: Process metrics.

@@

 Basili \& Boehm comment that since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!t]  
\begin{center}
\includegraphics[angle=270,width=4in]{img/b81.pdf}
 \end{center}
 \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}
Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.  McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this effect, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a proven
fact.
For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

This paper calls into question all the above claims about the delayed
issue effect. We suggest that the delayed issue effect might have been an dominant
effect decades ago,  but {\em not}  for $21^{\mathit{st}}$
software development. 
 The delayed issue effect was first reported in 1981 in a era of punch card programming
and non-interactive environments~\cite{Boehm81}. In the 21$^\mathit{st}$ century, we  program in 
interactive environments with higher-level languages and better source code control
tools. Such tools allow for the faster refactoring of existing
code-- in which case, 
managing the changes required to fix (say) an incorrect requirements assumption
is far less    onerous    than before.

@@

@@

That said, there are many empirical  findings that 
raise doubts that   general laws of SE even exist:
\be
\item
Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
on the relation of OO measures to defects.  Those results
 directly  contradict some of the laws listed by 
Endres \& Rombach~\cite{endres03}.
\item
Ray et al.~\cite{ray2014lang} tested if   strongly typed languages
predict for better code quality. In  728 projects,
they found  only a modest benefit in strong typing
(and caution that that effect may actually be due to other  conflating factors).
\item
Fenton \& Neil~\cite{fenton00,fenton00b}   critique the truism that
``pre-release fault rates for software
are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
amongst others). For the systems described in~\cite{fenton97}, they
show that software modules that were highly fault-prone
prior to release revealed very few faults after release.
\item
Meyer claims that   object-oriented (OO) encapsulation will
reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
that debugging an OO program is many times harder and
longer than debugging a standard procedural program~\cite{hatton98}.
\item
A truism of visual programming is that ``visual
representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
evidence for this claim is hardly conclusive~\cite{me00v}. 
\item
Numerous recent {\em local learning} results compare single models
learned from all available data to multiple models learned from clusters within the data~\cite{betten14,yang11,yang13,minku13,me12d,me11m,betta12,posnett11}.
A repeated result in those studies is that the local models generated the better effort
and defect predictions (better median results,
lower variance in the predictions).
\ee

To be fair, 
SE is  not the only
field where practitioners hang on to persistent beliefs, even if the evidence
for those beliefs is not strong.
The medical profession applies  many practices based on studies
that have been disproved. For example,
a  recent article
in the Mayo Clinic Proceedings~\cite{prasad13} found  
146 medical practices based on studies 
in year $i$, but which were  reversed by subsequent trials within years $i+10$.
Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
Aschwanden warns that ``cognitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.

The cognitive issues that complicate medicine are also found in software engineering.
According to Passos et al.~\cite{passos11}, commercial developers
are all too willing to believe in general
truisms.
They  warn that developers
usually assume that the truisms they learn from a few past
projects are general to 
all their future projects. They comment ``past experiences were taken into account without 
much consideration for their context~\cite{passos11}.''  
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
  supposed software engineering    ``experts'' rarely use lessons
  from past projects to improve their future reasoning~\cite{jorgensen09}. 
 They note that
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).

@@

@@

Our survey collected data on software engineers' views of commonly held software engineering ``laws''.
One of the laws questioned is a specific form of our delayed issue effect:  ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}). We abbreviate this law (illustrated in \fig{b81}) as RqtsErr.

(Technical aside: we use the RqtsErr formulation since this issue typically needs no supportive explanatory
text. If we had asked respondents about our more general term ``delayed issue
effect'', we would have had to burden our respondents with extra explanations).

The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers solicited through the Mechanical Turk;
participants were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology. The  second survey was conducted with Program Committee members of the ESEC/FSE 2015 and ICSE 2014  conferences solicited via email.

The respondents answered questions on two scales: \newline
\textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree. A text box was provided to explain the answer. \newline
\textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were: 
-1 meaning ``I don't know''; and 0 meaning ``this law does not apply at all'';
 1 meaning ``applies in a very narrow range of projects'';  2 meaning
``rarely applies'';   3 meaning
``occasionally applies'';  
4 meaning ``very frequently applies''; and
 5 meaning ``always applies''.
Respondents were required to explain the applicability score in a text box.

In order to baseline the participant's answers, participants were presented with the RqtsErr law and others. For the purposes
of this paper, the nature of the other laws other than RqtsErr are not relevant-- we
only added them in as a way to calibrate responses to the RqtsErr question. All laws were drawn from \cite{glass02} and \cite{endres03}. 
The PC member survey contained an additional question on ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix'' (the Delayed Issue Effect). Responses were recorded using the agreement question Likert scale.

Summary statistics for the agreement and applicability scores for the RqtsErr and DelayedIssueEffect laws are presented in Figure~\ref{fig:survey_results}. Responses whose Applicability response was ''I don't know'' are omitted from analysis.

\begin{figure}[!ht] 
\scriptsize 
 \begin{center}
\begin{tabular}{l|c|c|c|c|c}
 &  & \multicolumn{2}{c|}{agreement} & \multicolumn{2}{c}{applicability} \\\cline{1-1} 
Practitioner survey  & N & med & mode & med & mode \\
\hline 
\textbf{Rqts errors are most expensive...} & 16 & 5 & 5 & 4 & 5 \\ 
Inspections can remove 90\% of defects & 18 & 4 & 5 & 4 & 5 \\
80-20 rule (defects to modules) & 12 & 4 & 5 & 4 & 5 \\
Most time is spent removing errors & 16 & 4 & 4 & 4 & 5 \\ 
Process maturity improves output & 17 & 4 & 4 & 4 & 4 \\ 
Missing reqts are hardest to fix & 17 & 4 & 4 & 4 & 4 \\
Reuse increases prod. and qual. & 16 & 4 & 4 & 4 & 4 \\
OO-programming reduces errors & 13 & 4 & 4 & 4 & 3 \\
Adding manpower to a late project & 15 & 4 & 4 & 4 & 4 \\
Smaller changes have higher error density & 14 & 3 & 3 & 3.5 & 5 \\
A developer is unsuited to test own code & 17 & 3 & 1 & 4 & 5\\\hline

Researcher survey \\\hline 
Process maturity improves output & 4 & 4 & 4 & 4 & 5 \\
\textbf{Rqts errors are most expensive...} & 30 & 4 & 4 & 4 & 4   \\ 
\textbf{DelayedIssueEffect} & 30 & 4 & 4 & -- & --  \\
Reuse increases prod. and qual. & 6 & 4 & 4 & 4 & 4 \\
80-20 rule (defects to modules) & 6 & 4 & 4 & 4 & 3 \\
Missing reqts are hardest to fix & 7 & 4 & 4 & 4 & 3 \\
OO-programming reduces errors & 6 & 4 & 4 & 3 & 4 \\
Inspections can remove 90\% of defects & 7 & 4 & 4 & 3 & 3 \\
Adding manpower to a late project & 4 & 3 & 4 & 4 & 3 \\
Most time is spent removing errors & 6 & 3 & 3 & 4 & 4 \\ 
Smaller changes have higher error density & 4 & 3 & -- & 4 & 4 \\
A developer is unsuited to test own code & 7 & 2 & 1 & 3 & 3
\end{tabular} 
 \end{center}
\caption{Agreement and applicability of SE axioms.}
\label{fig:survey_results}
\end{figure}

Both  practitioners and researchers  strongly believed in RqtsErr: In both sets of responses, RqtsErr received  scores higher than most
other laws. Further, in the case of practitioners, this law was rated
as the single most believed effect.

Caveat: in the free reponse texts, we note that the researchers who disagreed with the law generally asserted that requirements change can be expensive, but that the effect depends on the process used (e.g., agile vs. waterfall) and the adaptability of the system architecture.

Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. 
This results strongly support that the notion that the  delayed issue effect   is a widely-held belief in software engineering.

@@

One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
in the SE literature. For example,
we know that \fig{b81} has been presented to the gradaute SE class at North Carolina State University, without quarrel or critical comment, every year for the last decade.

Yet when we look at the literature, the evidence for
delayed issue effect is both very sparse and very old.
The first data on the difficulties of resolving delayed issues as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}). These studies are most often cited by secondary sources regarding the delayed issue effect. We note that it is unclear from the text in ~\cite{Daly77} and \cite{Boehm76} if cost is defined in terms of effort, or in actual cost (i.e., labor, materiel, travel, etc).

These studies suggest that the difficulty (in terms of effort) to find and fix an error monotonically increases with lifecycle phase.  In 1990, Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).

\begin{figure}[!b]
\begin{center}\includegraphics[width=4in]{img/boehm_cost-to-fix.png}\end{center}
 \caption{Historical cost-to-fix curve. From~\cite{Boehm81}.}\label{fig:cost-to-fix}
 \end{figure}

In the 40 years since these initial studies, few studies have explored the difficulty to resolve issues
as a function of lifecycle phase. 
Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects. Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but we note here that the raw data points were not provided (which makes confirming those numbers 
a difficult task).

This was a common theme in the literature reviewed for this paper-- i.e.  that  it was no longer possible to access
the data used to make prior conclusions.
As an example of this, \fig{steck} shows one 2004 survey that reports an 
exponential delayed issue  effect for 
requirements issues in eight case studies. Note that we cannot verify
{\em any} of those results since the links in the references of
that  survey are all broken (``page not found'' errors).

As to the research into agile methods, one goal of that approach
is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Relatively little empirical data exists on this point.Elssamadisy and Schalliol~\cite{Elssamadisy02} anecdotally report on the growing, high cost of rework in a 50 person, three-year, 500KLOEC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their 
exact figures.

% We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake. To our knowledge, our research represents the first large-scale study of phase delay.

 \begin{figure}
{\small
\begin{center}
\begin{tabular}{r|rrrr}
 case& \multicolumn{4}{c}{Phase Requirements Issue Found }  \\
 study               &Requirements & Design & Code&  Test\\\hline
1& 1 &3& 5& 37\\  
2& 1  &    & 10 & 40 \\ 
3& 1   &    & 10 &  40 \\
4&  1  &   5 &      & 50  \\
5&  1 &3& 7& 51 \\
6& 1& 5 &33 &75  \\
7&  1 & 20 & 45 & 250 \\  
\end{tabular}
\end{center}}
\caption{Cost to resolve requirements issues, relative to resolving  during requirements. From ~\cite{steck04}.}\label{fig:steck}
\end{figure}

\begin{figure}[!t]
\begin{center}
 \includegraphics[width=4in]{img/clutterbuck.png} 
 \end{center}
 \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
 \end{figure}

 All in all, the empirical results of this section seem insufficient to justify
 the strength of the beliefs documented in the previous section. Adding to our doubts are  several studies that report less-than exponential
 increase in the difficulties associated with making the changes associated with delayed
 issues. 
Clutterbuck et al.~\cite{Clutterbuck09} studied a 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects). Clutterbuck et al. found the cost of change to be relatively flat until the later phases, with much of the effort spent in analysis of the change requests~\cite{Clutterbuck09}. Note that in this study, the effort increased by only 60\% (see the start and end of the curve in Figure~\ref{fig:clutterbuck}).

Another example of less-than exponential explosion in the difficulty associated with delayed issues comes from  Royce~\cite{Royce98}. He
studied  a million-line, safety-critical missile defense system
(see  Figure~\ref{fig:royce}). Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly. Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle.

 Other examples that report the opposite of DIE are:
  \bi
  \item
  Boehm~\cite{Boehm80} reported a flatter growth rate for small, non-critical projects.
   \item 
  Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to find certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).

\ei

\begin{figure}
\begin{center}\includegraphics[width=4in]{img/Royce98.png}\end{center}
 \caption{Exception to the rule - The Royce study: cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
 \end{figure}

 One feature of the above results will become very important in the subsequent discussion. We note that that in all
the results that report a delayed issue effect, those effects have began to manifest by the end of that project.

\section{Delayed Issues not  Harder  to Resolve}

The above analysis motivates a more detailed look at the delayed issued effect.  
Accordingly, we examined 171 software projects conducted between 2006 and 2014. These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$).

Since 2000, the SEI has been teaching and coaching TSP teams. One of the authors (Nichols) has mentored software development teams and coaches around the world as they deploy TSP within their organizations since 2006.  The  most recent completions were in 2014.
The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
Several projects extended for multiple incremental development cycles. 
Median team size was 7 people, with a maximum of 40. See \fig{dist} for the total
effort seen in those projects.

As to problem domain,
many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.

An anonymized version of that data is available in the PROMISE repository at openscience.us/repo.
For confidentiality restrictions, we cannot offer 
further details on these projects.

\subsection{About TSP$\textsuperscript{SM}$}

TSP is a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University~\cite{tsp00}. TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$) developed at the SEI by Watts Humphrey ~\cite{tsp00}. The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI. The Software Engineering Institute (SEI) at Carnegie Mellon University explores methods for software process improvement.

%TSP helps developers via a set of measures that can be applied to managing tasks, quality, and schedule. Planning begins by quantifying goals, defining work practices, and estimating size and effort. Developers then use this information to make a detailed short term plan. As the developers perform project work, they use a tool such as the Process Dashboard to collect their time, effort, size, and schedule data. Every week, the team reviews their data to evaluate status, identify actual rates, determine if project goals for schedule, cost, and quality are being met. The team then uses this information to make necessary plan corrections. At the end of the project the coach and team perform a quantitative project post mortem.

Common features of TSP projects include {\em planning}, {\em personal reviews}, {\em peer inspections}, and {\em coaching}.
A TSP {\em coach} helps the team to plan and analyze performance. The coach is the only role authorized to submit project data to the SEI.
Before reviewing data with the teams, therefore before submission, these coaches check the data for obvious errors.

During {\em Planning}, developers estimate the size of work products and convert this to a total effort using historical rates. Time in specific tasks come from the  process phases and historical percent time in phase distributions. Defects are estimated using historical phase injection rates and phase removal yields. Coaches help the developers to compare estimates against actual results. In this way, developers acquire a more realistic understanding of their work behavior, performance, and schedule status.

{\em Personal review} is a technique taken from the PSP and its use in TSP is unique.  Developers follow a systematic process to remove defects by  examining their own work products using a checklist built from their personal defect profile. This personal review occurs after some product or part of a product is considered to be constructed and before peer reviews or test.

%PSPSM teaches developers howto continually make and review their personnel estimates
%about their day-to-day tasks, then compare those estimates against the actual development effort.
%In this way, developers can acquire a more realistic understanding of their work behaviour.

\begin{figure}
\begin{center} 
\includegraphics[width=2.5in]{dom.pdf}
\end{center} 
\caption{Distribution of 
{\em effort} (which is
{\em team size} times
{\em days of work}). Median
value = 271 days.}\label{fig:dist}
\end{figure}

{\em Peer inspection} is a  technique in
traditional software engineering and is often called peer review.
 Basili and Boehm   commented in 2001~\cite{boehm01} 
that peer reviews can catch over half the defects introduced into a system.
Peer inspection can be conducted on any artifact generated anywhere in the software
lifecycle and can quickly be adapted to new kinds of artifacts. TSP peer reviews follow the Fagan style in which the reviewer uses a checklist composed of common team defects prior to a review team meeting.

Overall, the   effort associated with adding TSP to a project is not onerous. McHale reports~\cite{mchale02}:
\bi
\item
 The time spent  tracking time, defects, and tasks requires less then 3\% of a developer's time. Weekly team meetings  require at most an hour, which is
only 2.5\% of a 40 hour work week. 
\item
Team launches and replans average about 1 day per month or 5\% planning overhead.
\ei
It is true that one staff member is needed as a ``coach'' to mentor the teams
and certify and monitor that data collection. However, one of us (Nichols) has worked with dozens of TSP teams. He reports that one  trained coach can support 4 or 6 teams (depending upon team experience).

\subsection{Data Integrity}

A common property of real-world data sets is the presence
of noisy entries (superfluous  or spurious data). 
The level of noise can be quite high. As reported
in \cite{shepperd12}, around
10\% to 30\%
of the records in the NASA MDP defect data sets are
affected by noise.

One reason to use the SEI data for the analysis of this paper is its remarkably low level of noise.
Nichols et al.~\cite{shirai14}  report that
the noise levels in the SEI TSP data are smaller than those seen
in other data sets. They found in the SEI TSP data that:\bi 
\item
4\% of the data was incorrect (e.g. nulls, illegal formats);
\item  2\% of the data has inconsistencies such as timestamps
where the stop time was before the start time;
\item 3\% of the data contained values that were not credible
such as tasks listed in one day that took more than six hours for a single developer.
\ei 
One explanation for this low level of noise is the TSP process.
One the guiding principles of TSP was that  people performing the work are  responsible for planning and tracking the work. That is,  all the data collected here was entered
by local developers. This data was then checked by local coaches before being sent to the SEI
databases. Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data.
The use of certified local coaches within each project increases the integrity of our data.

\subsection{Data Details}
\label{sec:data-collection}

%Organizations using TSP agree to provide their project data to the SEI for use in research. In return the SEI agrees    that  data must not be traceable to its source. The data is collected at major project events; launch, interim checkpoints, and at project completion. Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the project post mortem report.  In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who must submit data to maintain their authorization. Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data not by the actual project results.  Of the data submitted, only the data recorded using the Process Dashboard tool has been collected and aggregated for this research. 
 %The key views include a project results summary, effort logs, task logs, and defect logs. The data has not yet undergone additional screening. A summary of the data quality issues was reported in a previous work ~\cite{shirai14} (that summary is discussed further in our {\em Validity} section, below).

Using tools provided by the SEI, developers kept very detailed logs of their daily activity. Our data includes  work start time, work end time,  delta
work time, and interruption time. Software engineers are often
interrupted by meetings, requests for technical help, reporting, and
so forth. These events are recorded, in minutes, as interruption
time.  
In this paper, when we report ``time to resolve an
issue,'' we show the difference between the start and end times
of a work session, with any interruption time subtracted (the
difference in times, minus the interruptions).
